
---
title: |
  | Barcelona Summer School of Demography
  | \vspace{1.5cm} \LARGE \emph{Module~2.~Demography with R}
  | \vspace{0.3cm} \huge \textbf{3.~Standardization and decomposition}\vspace{0.6cm}
fontsize: 11pt
geometry: a4paper, twoside, left=2.5cm, right=2.5cm, top=3.2cm, bottom=2.8cm, headsep
  = 1.35cm, footskip = 1.6cm
output:
  pdf_document:
    number_sections: yes
  html_document2: default
  html_document:
    number_sections: yes
    toc: yes
  pdf_document2: default
  header-includes:
    - \usepackage{titling}
    - \pretitle{\begin{center}\includegraphics[trim=0 0 0 8cm, width=6cm]{logotipCED.png}\\[\bigskipamount]}
    - \posttitle{\end{center}}
    - \usepackage{fancyhdr}
    - \pagestyle{fancy}
    - \fancyhead[LE]{\thepage~\qquad~Barcelona Summer School of Demography}
    - \fancyhead[RE]{Module~2.~Demography with R}
    - \fancyhead[LO]{Population growth}
    - \fancyhead[RO]{Tim Riffe\qquad~\thepage}
    - \fancyfoot[CO,CE]{\includegraphics[width=2.8cm]{logotipCED.png}}
subtitle: Population growth
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\noindent\makebox[\textwidth][c]{
  \begin{minipage}[t]{0.45\textwidth}
    \centering
    \Large{Tim Riffe} \\
    \vspace{0.1cm}\large{\texttt{tim.riffe@gmail.com}}
  \end{minipage}
}


\vspace{0.8cm}
\begin{center}
\large{8 July 2021}
\end{center}
\vspace{0.8cm}


\tableofcontents


# R functions
`sweep()`: Return an array obtained from an input array by sweeping out a summary statistic.

```{r error=F, message=F, warning=F}
#Example

Mat<-matrix(1:6, nrow=2, ncol=3)
vec1<-c(1:2)
vec2<-c(1:3)

sweep(Mat, 1, vec1)
sweep(Mat, 2, vec2)


```



# Matrix algebra: a brief review

```{r}
#Review of matrix algebra

#matrix multiplication

# Only possible if 
#the number of columns on the left matrix = 
#the number of rows on the right matrix

#Example 1
x<- matrix(c(1:9), ncol=3)
x

y<- matrix(c(1:6), ncol=2)
y

A<-x %*% y
A

A[1,1]
sum(x[1,]*y[,1])
#Example 2
x<- matrix(c(1,2,3), ncol=1)
x

y<- matrix(c(2,4,6), nrow=1)
y

A<-x %*% y
A


#matrix addition

B<-matrix(c(1:9), ncol=3)
C<-matrix(c(2:10), ncol=3)

D<-A+B
D[1,1]
A[1,1]+B[1,1]

```


# Data

We will use the Spanish data from the HMD [@HMD] and HFD [@HFD] for section 2 and 3 of this class. Please load the following:

```{r error=F, message=F, warning=F, echo=F}


setwd("C:/Users/mpbergeron/Dropbox/Teaching/Barcelona Summer School/2019/Day 4")

```


```{r}

#setwd()

#Births in 2014

B2014<-read.table("BirthsSpain2014.txt", header=T)
B2014$Age<-as.character(B2014$Age)
B2014$Age[B2014$Age=="12-"]<-"12"
B2014$Age[B2014$Age=="55+"]<-"55"
B2014$Age<-as.numeric(B2014$Age)

#Deaths in 2014

D2014<-read.table("DeathsSpain2014.txt", header=T)
D2014$Age<-as.character(D2014$Age)
D2014$Age[D2014$Age=="110+"]<-"110"
D2014$Age<-as.numeric(D2014$Age)

#Population in 2014 and 2015

P2014<-read.table("PopulationSpain2014.txt", header=T)
P2014$Age<-as.character(P2014$Age)
P2014$Age[P2014$Age=="110+"]<-"110"
P2014$Age<-as.numeric(P2014$Age)

P2015<-read.table("PopulationSpain2015.txt", header=T)
P2015$Age<-as.character(P2015$Age)
P2015$Age[P2015$Age=="110+"]<-"110"
P2015$Age<-as.numeric(P2015$Age)


# 2014 Life table (calculated in class 2)

LT<-read.table("LTSpain2014.txt", header=T)

```


# Crude growth

## Balancing equation 

Changes in population size is affected by changes in the number of births, deaths and migration.

* Two ways on entry: Birth and Immigration
* Two ways of exit: Death and Emigration

As there are only two ways of entry and two ways of exit, changes in population size comes from changes in the magnitude of these flows:

$$
N(T) = N(0) + B[0,T]-D[0,T]+I[0,T]-E[0,T]
$$


where

* $N(T)$ and $N(0)$ are the number of persons alive at time $T$ and $0$, respectively
* $B[0,t]$ is the number of births between times 0 and T
* $D[0,t]$ is the number of deaths between times 0 and T
* $I[0,t]$ is the number of immigrations between times 0 and T
* $E[0,t]$ is the number of emigrations between times 0 and T



## Crude growth rate

The crude growth rate is the change in population size relative to the person-year in a given time interval (e.g. from year 0 to T).

Giving the balancing equation, the crude growth rate (CGR) of a population is equal to:

$$
\frac{N(T)-N(0)}{PY[0,T]} = \frac{B[0,T]}{PY[0,T]} - \frac{D[0,T]}{PY[0,T]} + \frac{I[0,T]}{PY[0,T]} - \frac{E[0,T]}{PY[0,T]}
$$

$$
CGR[0,T] = \underbrace{CBR[0,T] - CDR[0,T]}_{CRNI[0,T]} + \underbrace{CIR[0,T] - CER[0,T]}_{CRNM_[0,T]}
$$

* $CBR[0,T]$ is the crude birth rate
* $CDR[0,T]$ is the crude death rate
* $CIR[0,T]$ is the crude immigration rate
* $CER[0,T]$ is the crude emigration rate
* $CRNI[0,T]$ is the crude rate of natural increase
* $CRNM[0,T]$crude rate of net migration


The CGR in Spain between 2014 and 2015 is
```{r}

# Person year

PY<-sum(P2015$Total+P2014$Total)/2

# Crude growth rate

CGR<- (sum(P2015$Total)-sum(P2014$Total))/PY
CGR


```

The crude rate of natural increase (CRNI) results from births and deaths processes only:

```{r}

# Crude birth rate
  
CBR<-sum(B2014$Total)/PY
CBR


# Crude death rate

CDR<-sum(D2014$Total)/PY
CDR


# Crude rate of natural increase

CRNI<- CBR - CDR
CRNI

```

The CGR suggests a decline of Spain's population size between 2014 and 2015. However, Spain has recorded a positive CRNI. A possible explanation is that there was more out-migration than in-migration in Spain in 2014, reducing the population size and opposing the natural increase. 



## Geometric growth

Different growth rates can be assumed in a population over time. If we consider a geometric growth over time, the annual rate of change (r) in a population at time 0 and t is equal to:

$$
r=\frac{N(T)-N(0)}{N(0)} \frac{1}{T}
$$

Thus, if r is constant over time the population size at time t can be found by

$$
N(t)=N(0)[1+r]^t
$$

```{r}
# If we assume that the rate of change observed in 2014 
#in Spain stays constant in the future, we have:


s.year<-2014
n.year<-c(0:10)

r.geom<-(sum(P2015$Total)-sum(P2014$Total))/sum(P2014$Total)

pop.geom<-sum(P2014$Total)*(1+r.geom)^n.year


#It we assume a close population (no migration)

r.geom2<-(sum(B2014$Total)-sum(D2014$Total))/sum(P2014$Total)

pop.geom2<-sum(P2014$Total)*(1+r.geom2)^n.year

plot(s.year+n.year, pop.geom, type="l",
     ylab="Population count", xlab="Year", ylim=c(min(pop.geom), max(pop.geom2)),
     main="Spanish population assuming geometric growth")
lines(s.year+n.year, pop.geom2, col=2)
legend("topleft", c("With migration", "Without migration"), 
       col=c(1,2), lty=1, bty="n")





```


## Exponential growth

The exponential growth can simply be seen as a continuous version of the geometric growth. Here, r is the instantaneous growth rate. If we consider an exponential growth over time, the annual rate of change (r) in a population at time 0 and t is equal to:

$$
r=\frac{ln \big(\frac{N(T)}{N(0)} \big)}{T}
$$
Thus, if r is constant over time the population size at time t can be found by

$$
N(t) = N(0)e^{r*t}
$$

```{r}

# Exponential growth

r.exp<-log(sum(P2015$Total)/sum(P2014$Total))

pop.exp<-sum(P2014$Total)*exp(r.exp*n.year)


#It we assume a close population (no migration)

r.exp2<-log((sum(P2014$Total)+sum(B2014$Total)-sum(D2014$Total))/sum(P2014$Total))

pop.exp2<-sum(P2014$Total)*exp(r.exp2*n.year)

plot(s.year+n.year, pop.exp, type="l",
     ylab="Population count", xlab="Year", ylim=c(min(pop.geom), max(pop.geom2)),
     main="Spanish population assuming exponential growth")
lines(s.year+n.year, pop.exp2, col=2)
legend("topleft", c("With migration", "Without migration"), 
       col=c(1,2), lty=1, bty="n")

```

## Doubling time

The doubling time is the number of year it would take for a population to double in size, under a fixed rate of change. In an exponential growth situation, if a population double between time 0 and T, then 

$$
ln \big( \frac{N(T)}{N(0)} \big) = ln(2)
$$

The doubling time (DB) for a population experiencing an exponential growth at rate r can then be found as:


$$
DB = \frac{ln(2)}{r}
$$

```{r}

# Doubling time

DB<-log(2)/r.exp
DB


# Doubling time without migration

DB2<-log(2)/r.exp2
DB2


```




# Fertility and replacement

## Replacement level

The replacement level is set at a TFR of 2.1. This is the average number of children a woman would need to have to reproduce herself by bearing a daughter who survives to childbearing age.

* TFR> 2.1: Population growth
* TFR= 2.1: Replacement level, constant population
* TFR< 2.1: Population decline

As the sex ratio at birth is generally 105 boys per 100 girls and because there is a risk of mortality, the TFR has to be greater than 2 for a woman to be able to "replace" herself by a daughter.



## Gross reproductive rate

The gross reproductive rate (GRR) is similar to the TFR but uses age-specific rate of having a female birth ($_nF^W_x[0,T]$)

$$
_nF^W_x[0,T] = \frac{Number~of~female~births~between~times~0~and~T~to~women~aged~x~to~x+n}{Number~of~person-years~lived~by~women~aged~x~to~x+n~between~times~0~and~T} 
$$

$$
_nF^W_x[0,T] \approx 0.488 * {}_nF_x
$$

$$
GRR[0,T] = n \sum_{x=a}^{B-n} {}_nF_x^W [0,T] \approx 0.488 *n \sum_{x=a}^{B-n} {}_nF_x [0,T]
$$


```{r}

F2014<-subset(P2014, Age %in% B2014$Age)[,c("Age", "Female")]
F2015<-subset(P2015, Age %in% B2014$Age)[,c("Age", "Female")]

# Person-years lived by women at reproductive ages
PY<-(F2014$Female+F2015$Female)/2


# ASFR
ASFR<-B2014$Total/PY

# TFR
TFR<-sum(ASFR)
TFR

# GRR
SR<-100/205
GRR<-sum(SR*ASFR)
GRR

```



## Net reproductive rate

The net reproductive rate (NRR) takes mortality into account. The NRR represents the average number of daughters that female in a cohort would bear in their lifespan if they were subject to the observed age-specific fertility ($_nF_x^W$) and mortality rates ($_nM_x$) [@preston2001demography]. The measure introduces the person-year lived between age x and x+n ($_nL_x$) in the GRR measure.


$$
NRR[0,T] = n \sum_{x=a}^{B-n} {}_nF_x^W [0,T] * \frac{_nL^W_x}{l_0}
$$


```{r}

LT.sub<-subset(LT, Age %in% B2014$Age)

PYx<-LT.sub$Lx/LT$lx[1]
NRR<- sum(SR*ASFR*PYx)
NRR


```


The NRR is a measure of growth!!!


# Population models

## Stationary vs Stable population

There are two main population models in demography: Stable and stationary. These models are theoretical types of population used to understand population dynamics under certain mortality and fertility assumptions.

### Stationary population

* Constant number of births
* Constant age-specific death rates
* Zero age-specific net-migration rates

Some properties

* Constant population size and structure
* CGR= 0
* $e_0 = \frac{1}{CBR} = \frac{1}{CDR}$


### Stable population

* Constant annual growth in birth: $B(t)=B(0) e^{rt}$
* Constant age-specific death rates
* Zero age-specific net-migration rates

The conditions are then:

* Constant age-specific fertility rates
* Constant age-specific death rates
* Age-specific net-migration rates are zero





## Leslie matrix 

The Leslie matrix is a matrix model used for population projections, assuming a stable population. The model combined mortality and fertility measures you have seen so far. It is a way to assess the long-term consequences, for the size of a population, of keeping a defined set of age-specific fertility and mortality rates constant over time. 

It takes the form of

\includegraphics[trim=7 12 7 3cm, clip=TRUE, width=1.1\textwidth]{Leslie.pdf}

where F* are adjusted fertility rates for mortality (see below) and s are the survivorship ratio from age x to x+n.

The Leslie matrix is, generally, conceive to forecast a close (no migration) female-only population.  

```{r}

#First, create a 0 matrix (111x111)

nr<-nrow(LT)
mat<-matrix(0, nr, nr)


```

### The cohort component method

The Leslie matrix is a matrix representation of the cohort component method. The matrix (see above) is filled in two steps.


**Step 1**: Project forward the women surviving age-category [x:x+n]

This step consists in finding the survivorship ratio ($s$) to each age group. The survivorship ratio is the proportion of people age x-n to x year that will be alive n years later, in a stationary population. 

$$
s_{x} = \frac{_nL_x}{_nL_{x-n}}
$$

```{r}

#Calculate survivorship ratio

surv<-LT$Lx[2:nr]/LT$Lx[1:(nr-1)]

```

The survivor at t+n time will thus be the people surviving from age x at time t to age x+n:

$$
_nN_x^W(t+n) = _nN_{x-n}^W(t) * s_x
$$
To represent this dynamic in the matrix, place the survivorship ratio as a diagonal in the matrix.

```{r}

#Place in matrix as diagonal

mat[row(mat)==col(mat)+1] <- surv

```

For the open age interval, the population at time t+n is equal to

$$
_{\infty}N_x^W(t+n) = _nN_{x-n}^W(t) * s_x + _{\infty}N_{x}^W(t) *\frac{T_{x+n}}{T_x}
$$
```{r}

#The last cell

mat[nr,nr]<-(LT$Tx[nr]/LT$Tx[nr-1])

```


**Step 2**: Finding the number of surviving female in the first age group (age 0)

This step consists in finding the number of female births during the projected period, which survive until the end of the period. 

$$
_nN_0^W(t+n) = B^W[t,t+n]*\frac{_nL_0}{n*l_0}
$$
The way to calculate the number of births is:

$$
B[t,t+n] = \sum_{x=\alpha}^{\beta-n} {}_nF_x * n * \frac{(_nN_x^W (t) +  _nN_{x}^W(t+n))}{2} 
$$
$$
 = \sum_{x=\alpha}^{\beta-n} {}_nF_x * n * \frac{(_nN_x^W (t) +  _nN_{x-n}^W(t) \frac{_nL_x}{_nL_{x-n}})}{2} 
$$
Thus, $_nN_0^W(t+n)$ is equal to:

$$
_nN_0^W(t+n) =\frac{_nL_0}{2*l_0} \frac{1}{1+SRB} \sum_{x=\alpha}^{\beta-n} {}_nF_x * n * \big(_nN_x^W (t) +  _nN_{x-n}^W(t) \frac{_nL_x}{_nL_{x-n}}\big) 
$$
where SRB is the sex ratio at birth.

```{r}

#Create a vector
ASFR2<-rep(0, nr)
ASFR2[B2014$Age+1]<-ASFR

#Constant across age of mother in N(0) equation
const<- (1/(1+1.05))*(LT$Lx[2]/(2*LT$lx[1]))

#Non-constant across mother age
firstrow <- const*(ASFR2[1:(nr-1)]+(ASFR2[2:nr]*surv[1:(nr-1)]))

#Fill the first row
mat[1,] <- c(firstrow,0)

```

### Projections
To forecast the population at time t (P(t)), for 1-year time interval, then

$$P(t)=L^t * P(0)$$

where L is the Leslie matrix.


```{r error=F, message=F, warning=F}

#Population at time 0

P0<-matrix(P2015$Female, ncol=1)


#Forecast
time <- c(0:300)

projection <- function(mat, time, pop){
  for(i in 1:time){
    if(i==1){project<-mat
    proj <- project%*%pop}
    if(i>1){project<- project%*% mat
    proj<-project%*%pop}
  }
  return(proj)
}

proj.pop<- matrix(0, nrow(mat), length(time))
for (i in 1:ncol(proj.pop)){
  proj.pop[,i]<-projection(mat=mat, time=time[i], pop=P0)
}


#Total population
tot.pop<-colSums(proj.pop)

plot(x=2015+time, y=tot.pop, type="l",
     ylab="Population count", xlab="Year",
     main="Population size for Spanish female forecast")


#Number of births

birth<-proj.pop[1,]

plot(x=2015+time, y=birth, type="l",
     ylab="Number of births", xlab="Year",
     main="Number of births for Spanish female forecast")

library(ggplot2)
Age<-c(0:110)

H<-10
ggplot() + 
  geom_bar(aes(x = Age, y = proj.pop[,H]), stat = "identity", fill="red") +
  ylim(c(0, max(proj.pop)))+
  ylab("Population counts")+
  coord_flip()+
  ggtitle(paste("Population counts forecast in", 2015+H))+
  theme_bw()


H<-100
ggplot() + 
  geom_bar(aes(x = Age, y = proj.pop[,H]), stat = "identity", fill="red") +
  ylim(c(0, max(proj.pop)))+
  ylab("Population counts")+
  coord_flip()+
  ggtitle(paste("Population counts forecast in", 2015+H))+
  theme_bw()

```

## Leslie matrix and stable population properties

### Constant population structure and crude growth rate

On the long term, the population age structure converges towards a constant age-structure.

```{r}

perc.pop<-sweep(proj.pop, 2, tot.pop, FUN="/")

plot(x=2015+time, perc.pop[1,], type="l", ylim=c(0, 0.02),
     ylab="Proportion", xlab="Year",
     main="Proportion of Spanish female population at specific ages")
lines(x=2015+time, perc.pop[26,], col=2)
lines(x=2015+time, perc.pop[46,], col=3)
lines(x=2015+time, perc.pop[66,], col=4)
lines(x=2015+time, perc.pop[86,], col=6)
legend("bottomleft", c("0", "25", "45", "65", "85"),
       lty=1, col=c(1:4,6), bty="n", cex=0.7)
```


The crude growth rate also converges towards a constant over time in a stable population.

```{r}

CGR.proj<-c(1:(length(time)-1))
for(i in 1:length(CGR.proj)){
  CGR.proj[i]<-log(sum(proj.pop[,i+1])/sum(proj.pop[,i]))
}

plot(2015+time[1:(length(time)-1)], CGR.proj, type="l",
     ylab="CGR", xlab="Year",
     main="Crude growth rate of Spanish female population")

```



# Exercises {-}

1) Load 'LTFrance2015.txt', 'ASFRFrance2015.txt' and 'PopFrance2015.txt'.

2) Calculate the TFR and NRR for France in 2015. Discuss if France reached the population replacement level in 2015.

3) Create the Leslie matrix for French females in 2015.

4) Project French females population 200 years ahead. What is the proportion of the 2015 population left 200 later?

5) How did the age-structure change between 2015 and 2215? Older or younger?

6) Calculate the CGR for France. Discuss your results in relation with the CGR for Spain found in the class. 



# Eigenvalue as crude growth rate in stable populations {-}

Eigen comes from the German "self". The eigenvalues can be referred to as "latent roots" of a squared matrix. They play a fundamental role in the solution of linear system and provide concrete information "*from a set of static, algebric equations*" [@caswell2001matrix].

The eigenvalue ($\lambda$) of a matrix A is that

$$
Ax = \lambda x
$$
where x is some scalar, called eigenvector.

The eigenvalue is found by solving

$$
(A-\lambda I)x = 0
$$
where I is the identity matrix.

For example (issued from @caswell2001matrix)

```{r}

#Example

A<-matrix(c(3,2,-6,-5), ncol=2)
A

#if x=t(1,1)
x<-matrix(c(1,1), ncol=1)
x

#then lambda
lambda<- A%*%x

#x is thus the eigenvector of A and lambda is -3
lambda


```

The intrinsic crude growth rate of a stable population is found by finding the eigenvalue of the Leslie matrix, $CGR = 1-\lambda$.

```{r}

-(1-as.numeric(eigen(mat)$values[1]))
CGR.proj[length(time)-1]

```

The intrinsic population structure is the first eigenvector of the Leslie matrix.

```{r}

prop<-eigen(mat)$vector[,1]/sum(eigen(mat)$vector[,1])


plot(LT$Age, as.numeric(prop),
     ylab="Age", xlab="Proportion",
     main="Intrinsic population structure for stable population")
lines(perc.pop[,300], col=2)

```



# References {-}
