---
title: "class notes"
author: "Tim Riffe"
date: "7/9/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary


# Data
```{r, include = FALSE}
library(tidyverse)
library(demography)
data  <- hmd.mx("ESP", us, pw)
sexes <- data$pop %>% names()

# two containers, columns given, but no rows
ESpop   <- tibble(Year = NULL, Age = NULL, Sex = NULL, Exposure = NULL)
ESrates <- tibble(Year = NULL, Age = NULL, Sex = NULL, M = NULL)

for (i in sexes){
  ESpop <- data$pop[[i]] %>% 
    as_tibble() %>% 
    rownames_to_column("Age") %>% 
    pivot_longer(cols = -Age,     
                 names_to = "Year", 
                 values_to = "Exposure") %>% 
    mutate(Sex = i,
           Age = as.integer(Age) - 1) %>% 
    bind_rows(ESpop)
  
  ESrates  <- data$rate[[i]] %>% 
    as_tibble() %>% 
    rownames_to_column("Age") %>% 
    pivot_longer(cols = -Age, names_to = "Year", values_to = "M") %>% 
    mutate(Sex = i,
           Age = as.integer(Age) - 1) %>% 
    bind_rows(ESrates)
}

ES <- left_join(ESpop, 
                ESrates,
                by = c("Age", "Year", "Sex")) %>% 
  select(Year, Sex, Age, Exposure, M) %>% 
  arrange(Year, Sex, Age) %>% 
  mutate(Year = as.integer(Year)) %>% 
  filter(Year >= 1950,
         Year <= 2014,
         Sex == "female") %>% 
  # get Deaths
  # then group to ages 100+
  mutate(Deaths = M * Exposure,
         Age = ifelse(Age >= 100, 100, Age)) %>% 
  group_by(Year, Sex, Age) %>% 
  summarize(Exposure = sum(Exposure),
            Deaths = sum(Deaths, na.rm = TRUE),
            .groups = "drop") %>% 
  # Finally get back Mx
  mutate(M = Deaths / Exposure)
```


# The Lee-Carter model

$$
ln(m_{x}(t)) = \alpha_x + \beta_x \kappa(t) + \epsilon_x(t)
$$
where $\alpha_x$ is the age-specific average of the log death rates ($m_x(t)$). The $\beta_x$ and $\kappa_t$ are found with a Singular Value Decomposition (SVD) and consist in the normalized first right and left singular vectors. They are interpreted as the age-pattern (age-specific rate of change relative to $\kappa(t)$) of mortality change and the time-pattern of mortality change. Mortality is forecast by extrapolating $\kappa(t)$. The term $\epsilon_x(t)$ is the error of the model.

### check for log-linearity

The Lee-Carter approach is premised on gradual log-linear change in mortality rates
```{r error=F, message=F, warning=F}

ES %>% 
  filter(Age %% 10 == 0) %>% 
  ggplot(aes(x = Year, y = M, color = Age, group = Age)) +
  geom_line() +
  scale_y_log10()
```

Let's go step by step!

## Calculate the age-specific average of the log death rates. 
```{r error = F, message = F, warning = F}
#Log transformed data
lM <-
  ES %>% 
  select(Year, Age, M) %>% 
  pivot_wider(names_from = Year, values_from = M) %>% 
  column_to_rownames("Age") %>% 
  as.matrix() %>% 
  log()

ax  <- rowMeans(lM)
Age <- rownames(lM) %>% as.integer()
```


Examine $\alpha_x$ in a base-tastic plot
```{r error=F, message=F, warning=F}
plot(Age, ax, type="l",
     ylab=expression(alpha[x]), xlab="Age",
     main=expression("Age-specific average"~alpha[x]~", Spanish females 1950-2014"))

```


## Center the matrix on $\alpha_x$

```{r error=F, message=F, warning=F}

#Center the matrix
cent_lM <- sweep(lM, 1, ax, FUN = "-")

```

## Do SVD on the centered matrix

We now do singular value decomposition on the centered matrix of log mortality. Here, the left-singular vectors capture the age-pattern of mortality change and the right-singular vectors, the time-pattern of mortality change. For further insight on SVD in demography, take a look at Monica Alexander's 2017 blog post [https://www.monicaalexander.com/posts/2017-12-16-svd/](https://www.monicaalexander.com/posts/2017-12-16-svd/)


```{r error=F, message=F, warning=F}
#SVD: using first vectors and value only

svd_lM <- svd(cent_lM)

u      <- svd_lM$u[, 1]
v      <- svd_lM$v[, 1]
d      <- svd_lM$d[1]

#Explained variance

exp.var<- cumsum((svd_lM$d)^2 / sum((svd_lM$d)^2))[1]
exp.var
```

In order to reach a unique solution, the parameters are normalized such that $\sum_x \beta_x = 1$ and $\sum_t \kappa(t) = 0$. 

```{r error=F, message=F, warning=F}

#Normalization

Bx <- u / sum(u)
plot(Age, Bx, type="l",
     ylab=expression(beta[x]), xlab="Age",
     main=expression("Age-pattern of mortality change"~beta[x]~", Spanish females 1950-2014"))
abline(h=0, lty=2)

Kt<-v*sum(u)*d
plot(1950:2014, Kt, type="l",
     ylab=expression(kappa(t)), xlab="Age",
     main=expression("Time-pattern of mortality change"~kappa(t)~", Spanish females 1950-2014"))
abline(h=0, lty=2)
```


## Forecast $\kappa_t$

```{r error=F, message=F, warning=F}

#Forecast horizon
h      <- 1:100
Year   <- 1950:2014
#Drift
lK     <- length(Kt)
drift  <- (Kt[lK] - Kt[1]) / (lK - 1)
drift

#Forecast
Kt.fcst <- Kt[lK] + drift * h

plot(Year, Kt, type = "l", ylim=c(min(Kt.fcst), max(Kt)), 
     xlim = c(Year[1], Year[length(Year)]+h[length(h)]),
     ylab = expression(kappa(t)), xlab="Age",
     main = expression(~kappa(t)~"observed and forecast, Spanish females 1950-2014"))
lines(Year[length(Year)]+h, Kt.fcst, col=2)

```



## Find the prediction intervals (PI) 

The prediction intervals (PI) are found by finding the standard errors of the estimates (SE) from the random walk with drift model. The SE indicates the uncertainty with one year forecast ahead. It is here assumed that the SE increase with the forecast horizon (h) square root.

$$
PI(t+h) = \kappa(t+h) +/- z_{1-a} SE \sqrt{h}
$$
where $z$ is the z-score at level $a$
```{r error=F, message=F, warning=F}

# Fit random walk with drift
fit.Kt <- Kt[1:(lK - 1)] + drift

# Find SE
se     <- sqrt(sum((Kt[2:lK] - fit.Kt)^2) / (length(Kt) - 1))

# Calculate PI
Kt.95u <- Kt.fcst + 1.96 * se * sqrt(h)
Kt.95l <- Kt.fcst - 1.96 * se * sqrt(h)


plot(Year, Kt, type="l", ylim=c(min(Kt.95l), max(Kt)), 
     xlim=c(Year[1], Year[length(Year)]+h[length(h)]),
     ylab=expression(kappa(t)), xlab="Age",
     main=expression(~kappa(t)~"observed and forecast, Spanish females 1950-2014"))
lines(Year[length(Year)]+h, Kt.fcst, col=2)
lines(Year[length(Year)]+h, Kt.95u, col=2, lty=2)
lines(Year[length(Year)]+h, Kt.95l, col=2, lty=2)
```


**Step 6**: Reconstruct and back-transform the matrix

This is just taking the reverse steps of what we first did to $M_x$.
```{r error=F, message=F, warning=F}
Bx         <- matrix(Bx, ncol = 1)
mx.fit     <- exp(Bx %*% Kt + ax) 
mx.fcst    <- exp(Bx %*% Kt.fcst + ax)
mx.fcst95u <- exp(Bx %*% Kt.95u + ax)
mx.fcst95l <- exp(Bx %*% Kt.95l + ax)
```

**Step 8** Let's grab all those pieces and make it tidy again for further calculations and processing.

```{r}
# add dimension names
dimnames(mx.fit)     <- dimnames(lM)
dimnames(mx.fcst)    <- list(Age = 0:100, Year = 2014 + h)
dimnames(mx.fcst95u) <- list(Age = 0:100, Year = 2014 + h)
dimnames(mx.fcst95l) <- list(Age = 0:100, Year = 2014 + h)

# longer
mx_fit <-
  mx.fit %>% 
  as_tibble(rownames = "Age") %>% 
  pivot_longer(-Age,
               names_to = "Year",
               values_to = "M") %>% 
  mutate(variant = "fitted")

mx_fcst <-
  mx.fcst %>% 
  as_tibble(rownames = "Age") %>% 
  pivot_longer(-Age,
               names_to = "Year",
               values_to = "forecast")
mx_fcst95u <-
  mx.fcst95u %>% 
  as_tibble(rownames = "Age") %>% 
  pivot_longer(-Age,
               names_to = "Year",
               values_to = "lower")
mx_fcst95l <-
  mx.fcst95l %>% 
  as_tibble(rownames = "Age") %>% 
  pivot_longer(-Age,
               names_to = "Year",
               values_to = "upper")

# bind the pieces together into single object
fcst <- mx_fcst %>% 
  left_join(mx_fcst95u, by = c("Age", "Year")) %>% 
  left_join(mx_fcst95l, by = c("Age", "Year")) %>% 
  pivot_longer(forecast:upper, 
               names_to = "variant",
               values_to = "M") %>% 
  bind_rows(mx_fit) %>% 
  mutate(Age = as.integer(Age),
         Year = as.integer(Year))

```

**Step 7**: Calculate the life table

```{r error=F, message=F, warning=F}
LTfcst <- 
  fcst %>% 
  group_by(Year, variant) %>% 
  mutate(M = ifelse(is.na(M), .5, M),         # hack
         n = 1,
         ax = case_when(
                Age == 0 & M < .02012 ~ .14916 - 2.02536 * M,
                Age == 0 & M < .07599 ~ 0.037495 + 3.57055 * M,
                Age == 0 & M >= .07599 ~ 0.30663,
                Age == 110 ~ 1 / M,
                TRUE ~ n / 2),
          ax = ifelse(is.infinite(ax),.5,ax),
          qx = (M * n) / (1 + (n - ax) * M),
          qx = ifelse(qx > 1, 1, qx),
          px = 1 - qx,
          lx = c(1, cumprod(px[-n()])),
          dx = qx * lx,
          Lx = lx - (n - ax) * dx,
          Tx = Lx %>% rev() %>% cumsum() %>% rev(),
          ex = Tx / lx)

```

## Visualize results

Life expectancy at birth
```{r}
LTfcst %>% 
  filter(Age == 0) %>% 
  ggplot(aes(x = Year, y = ex, group = variant)) +
  geom_line()
```

Life expectancy at age 65
```{r}
LTfcst %>% 
  filter(Age == 65) %>% 
  ggplot(aes(x = Year, y = ex, group = variant)) +
  geom_line()
```


